---
title: "King County Food Establishment Safety"
format: dashboard
embed-resources: true
---

```{r setup}
#| include: false
library(ggplot2)
library(tidyverse)
library(leaflet)
library(DT)

inspection <- read.csv("Food_Establishment_Inspection_Data.csv", stringsAsFactors = FALSE)

# Data Cleaning 

inspection_clean <- inspection |>
  mutate(
    Address          = str_trim(str_to_upper(Address)),
    Name             = str_trim(str_to_upper(Name)),
    City_norm        = str_trim(str_to_upper(City)),

    Inspection.Score = as.numeric(Inspection.Score),
    Inspection.Date  = as.Date(Inspection.Date, "%m/%d/%Y"),
    Year             = year(Inspection.Date),

    unsatisfactory = Inspection.Result == "Unsatisfactory",

    # Collapse the many result values into four groups
    result_group = case_when(
      Inspection.Result == "Satisfactory"   ~ "Satisfactory",
      Inspection.Result == "Unsatisfactory" ~ "Unsatisfactory",
      str_detect(Inspection.Result, "Not|Incomplete|No Longer") ~ "Inspection Issues",
      TRUE ~ "Administrative"
    )
  )

# Keep only rows with a clear pass/fail
inspection_valid <- inspection_clean |>
  filter(result_group %in% c("Satisfactory", "Unsatisfactory"))

summary_stats <- inspection_clean |>
  summarise(
    total_inspections   = n(),
    total_businesses    = n_distinct(Business_ID),
    avg_score           = mean(Inspection.Score, na.rm = TRUE),
    unsatisfactory_rate = mean(unsatisfactory, na.rm = TRUE)
  )

result_counts <- inspection_clean |>
  count(result_group)
```

# Dataset Overview

This dashboard explores the King County Food Establishment Inspection dataset.  
Each row of the original dataset represents an inspection of a food establishment, 
including its score, result (Satisfactory/Unsatisfactory), location, and any 
violations recorded.

We aim to understand:

* How often inspections are unsatisfactory
* Where in King County food safety issues are concentrated
* What factors most commonly contribute to high-risk and low-risk health violations

A food safety rating of 0 is best. The classification of "Excellent", "Good",
and "Okay" depends on both the rating, the Food Service Establishment Risk Level,
and the timeframe of inspections being taken into account. An in-depth explanation of what goes into a restaurant's
rating can be found at the [King County Department of Public Health](https://kingcounty.gov/en/dept/dph/health-safety/food-safety/inspection-rating-system/rating-system).

## Row {height=15%}

```{r}
#| content: valuebox
#| title: "Total Inspections"
list(icon = "clipboard-check", color = "#0d6efd", value = format(summary_stats$total_inspections, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Unique Businesses"
list(icon = "shop", color = "#0dcaf0", value = format(summary_stats$total_businesses, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Overall Unsatisfactory Rate"
list(icon = "exclamation-triangle", color = "#ffc107",
     value = paste0(round(summary_stats$unsatisfactory_rate * 100, 1), "%"))
```

```{r}
#| content: valuebox
#| title: "Avg Inspection Score"
list(icon = "bar-chart", color = "#dc3545",
     value = round(summary_stats$avg_score, 1))
```

## Row {height=40%}

```{r}
#| title: "Distribution of Inspection Scores"
#| warning: false
#| message: false

# Cap at 100 to exclude extreme outliers
ggplot(inspection_clean |> filter(Inspection.Score >= 0, Inspection.Score <= 100),
       aes(x = Inspection.Score)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", linewidth = 0.3) +
  labs(x = "Inspection Score (0 = best)", y = "Count") +
  theme_minimal()
```

```{r}
#| title: "Inspection Outcomes"
#| warning: false
ggplot(result_counts, aes(x = reorder(result_group, n), y = n, fill = result_group)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Satisfactory"      = "#2ca02c",
    "Unsatisfactory"    = "#d62728",
    "Inspection Issues" = "#ff7f0e",
    "Administrative"    = "#aec7e8"
  )) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Number of Inspections") +
  theme_minimal()
```


# Where Problems Occur

This page explores the geographic distribution of food inspection results across King County.
Each point represents an individual inspection record — **red = unsatisfactory, green = satisfactory**.

```{r}
#| warning: false
#| message: false
#| title: "King County Inspection Outcomes Map"

set.seed(42)
map_data <- inspection_valid |>
  filter(
    !is.na(Latitude), !is.na(Longitude),
    # Filters out erroneous coordinates outside King County
    Latitude  > 47.0, Latitude  < 47.9,
    Longitude > -122.6, Longitude < -121.7
  ) |>
  slice_sample(n = 15000) # Sample to 15,000 rows for better visibility

# Map true/false unsatisfactory value to red/green
pal <- colorFactor(c("#d62728", "#2ca02c"), domain = c(TRUE, FALSE))

leaflet(map_data) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = -122.15, lat = 47.5, zoom = 10) |>
  addCircleMarkers(
    lng         = ~Longitude,
    lat         = ~Latitude,
    color       = ~pal(unsatisfactory),
    radius      = 3,
    stroke      = FALSE,
    fillOpacity = 0.55,
    popup = ~paste0(
      "<b>", Name, "</b><br>",
      Address, ", ", City, "<br>",
      "Result: ", Inspection.Result, "<br>",
      "Score: ", Inspection.Score, "<br>",
      "Date: ", Inspection.Date
    )
  ) |>
  addLegend("bottomright",
            colors  = c("#d62728", "#2ca02c"),
            labels  = c("Unsatisfactory", "Passed"),
            title   = "Inspection Result",
            opacity = 0.8)
```


# Unsatisfactory Rates by City & ZIP

This page breaks down unsatisfactory result rates geographically, by city and by ZIP code. This identifies
communities in King County with the highest concentration of food safety issues.

## Row

```{r}
#| title: "Unsatisfactory Rate by City (≥ 200 inspections)"
#| warning: false

# Require at least 200 inspections per city to keep results meaningful
city_stats <- inspection_valid |>
  group_by(City_norm) |>
  summarise(
    n_inspections       = n(),
    unsatisfactory_rate = mean(unsatisfactory, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_inspections >= 200) |>
  arrange(desc(unsatisfactory_rate)) |>
  slice_head(n = 20)

ggplot(city_stats, aes(x = reorder(City_norm, unsatisfactory_rate),
                       y = unsatisfactory_rate, fill = unsatisfactory_rate)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.75)) +
  scale_fill_gradient(low = "#fee08b", high = "#d62728") +
  labs(x = NULL, y = "Unsatisfactory Rate") +
  theme_minimal()
```

```{r}
#| title: "ZIP Code Summary Table"
#| warning: false

# Interactive table. Users can sort and search across all King County ZIPs
zip_table <- inspection_valid |>
  mutate(Zip.Code = as.character(Zip.Code)) |>
  group_by(Zip.Code) |>
  summarise(
    Inspections           = n(),
    Unsatisfactory        = sum(unsatisfactory, na.rm = TRUE),
    `Unsatisfactory Rate` = paste0(round(mean(unsatisfactory, na.rm = TRUE) * 100, 1), "%"),
    `Avg Score`           = round(mean(Inspection.Score, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  filter(Inspections >= 50) |>   # drop ZIPs with too few records
  arrange(desc(Unsatisfactory))

datatable(zip_table,
          options = list(pageLength = 12, scrollX = TRUE),
          rownames = FALSE)
```

# Trends Over Time

How have inspection outcomes changed over the years? This page tracks unsatisfactory result rates 
and inspection volumes over time across all of King County.

## Row

```{r}
#| title: "Annual Unsatisfactory Rate Over Time"
#| warning: false

# Restrict to 2010-2024, earlier years have sparse coverage
annual <- inspection_valid |>
  filter(!is.na(Year), Year >= 2010, Year <= 2024) |>
  group_by(Year) |>
  summarise(
    n_inspections       = n(),
    unsatisfactory_rate = mean(unsatisfactory, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(annual, aes(x = Year, y = unsatisfactory_rate)) +
  geom_line(color = "#d62728", linewidth = 1.2) +
  geom_point(color = "#d62728", size = 2.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(2010, 2024, 2)) +
  labs(x = "Year", y = "Unsatisfactory Rate") +
  theme_minimal()
```

```{r}
#| title: "Annual Inspection Volume"
#| warning: false

# Uses the same data frame computed in the chunk above
ggplot(annual, aes(x = Year, y = n_inspections)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2010, 2024, 2)) +
  labs(x = "Year", y = "Number of Inspections") +
  theme_minimal()
```

## Row

```{r}
#| title: "Unsatisfactory Rate by Month (All Years)"
#| warning: false

# Aggregate across all years to surface any seasonal inspection patterns
monthly <- inspection_valid |>
  filter(!is.na(Inspection.Date)) |>
  mutate(Month = month(Inspection.Date, label = TRUE, abbr = TRUE)) |>
  group_by(Month) |>
  summarise(
    n_inspections       = n(),
    unsatisfactory_rate = mean(unsatisfactory, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(monthly, aes(x = Month, y = unsatisfactory_rate, group = 1)) +
  geom_line(color = "#ff7f0e", linewidth = 1.2) +
  geom_point(color = "#ff7f0e", size = 2.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Month", y = "Unsatisfactory Rate",
       caption = "Aggregated across all years in the dataset") +
  theme_minimal()
```

# Violation Type Breakdown

This page examines the proportion of individual **RED** (critical) vs **BLUE** (non-critical) violations. Per King County, Red violations directly cause food-borne illnesses (improper temperatures, poor hygiene for example), while blue violations cover maintenance and sanitation.

## Row {height=30%}

```{r}
#| content: valuebox
#| title: "RED Violations"

# Empty strings are present in Violation.Type for non-violation rows; == "RED" handles this safely
red_v <- sum(inspection_clean$Violation.Type == "RED", na.rm = TRUE)
list(icon = "exclamation-circle", color = "#d62728", value = format(red_v, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "BLUE Violations"
blue_v <- sum(inspection_clean$Violation.Type == "BLUE", na.rm = TRUE)
list(icon = "info-circle", color = "#1f77b4", value = format(blue_v, big.mark = ","))
```

## Row

```{r}
#| title: "RED vs BLUE Violation Proportion by Establishment Category"
#| warning: false

# Exclude empty strings, meaning rows with no violation cited
inspection_clean |>
  filter(!is.na(Violation.Type), Violation.Type != "", !is.na(Description)) |>
  mutate(Description = str_trunc(Description, 45)) |>
  group_by(Description, Violation.Type) |>
  summarise(count = n(), .groups = "drop") |>
  group_by(Description) |>
  mutate(
    total = sum(count),
    prop  = count / total   # proportion within each category
  ) |>
  ungroup() |>
  filter(total >= 500) |>   # drop categories with too few violations to be meaningful
  # reorder bars so highest RED proportion appears at top
  ggplot(aes(x = reorder(Description, prop * (Violation.Type == "RED")),
             y = prop, fill = Violation.Type)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("RED" = "#d62728", "BLUE" = "#1f77b4")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Proportion of Violations", fill = "Violation Type") +
  theme_minimal(base_size = 10)
```

```{r}
#| title: "RED vs BLUE Proportion by City (>= 200 violations)"
#| warning: false

# slice_max on total keeps only the highest volume cities to avoid reduce y-axis
inspection_clean |>
  filter(!is.na(Violation.Type), Violation.Type != "") |>
  group_by(City_norm, Violation.Type) |>
  summarise(count = n(), .groups = "drop") |>
  group_by(City_norm) |>
  mutate(
    total = sum(count),
    prop  = count / total
  ) |>
  ungroup() |>
  filter(total >= 200) |>
  slice_max(order_by = total, n = 40) |>
  ggplot(aes(x = reorder(City_norm, prop * (Violation.Type == "RED")),
             y = prop, fill = Violation.Type)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("RED" = "#d62728", "BLUE" = "#1f77b4")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Proportion of Violations", fill = "Violation Type") +
  theme_minimal(base_size = 10)
```
