---
title: "STAT-408 Final Project"
format: dashboard
---

# Dataset Overview

This dashboard explores the King County Food Establishment Inspection dataset.  
Each row of the original dataset represents an inspection of a food establishment, 
including its score, result (Satisfactory/Unsatisfactory), location, and any 
violations recorded.

We aim to understand:

* How often inspections fail
* Where in King County food safety issues are concentrated
* What factors are associated with poor inspection outcomes

A food safety rating of 0 is best. The classification of "Excellent", "Good",
and "Okay" depends on both the rating, the Food Service Establishment Risk Level,
and the timeframe of inspections being taken into account. An in-depth explanation of what goes into a restaurant's
rating can be found at the official King County department of public health 
website found [here](https://kingcounty.gov/en/dept/dph/health-safety/food-safety/inspection-rating-system/rating-system
)

## Row {height=40%}
```{r}
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 16

library(ggplot2)
library(tidyverse)

inspection <- read.csv("Food_Establishment_Inspection_Data.csv")

# Cleaning
inspection_clean <- inspection |>
  mutate(
    # Normalize text fields 
    Address = str_trim(str_to_upper(Address)),
    Name = str_trim(str_to_upper(Name)),
    
    Inspection.Score = as.numeric(Inspection.Score),
    Inspection.Date = as.Date(Inspection.Date, "%m/%d/%Y"),
    
    # Binary variable for inspection which failed
    failed = Inspection.Result == "Unsatisfactory"
  )

# Group inspection process issues
inspection_clean <- inspection_clean |>
  mutate(
    result_group = case_when(
      Inspection.Result == "Satisfactory" ~ "Satisfactory",
      Inspection.Result == "Unsatisfactory" ~ "Unsatisfactory",
      str_detect(Inspection.Result, "Not|Incomplete|No Longer") ~ "Inspection Issues",
      TRUE ~ "Administrative"
    )
  )

# Compute overall summary statistics
summary_stats <- inspection_clean |>
  summarise(
    total_inspections = n(),
    total_businesses = n_distinct(Address),
    avg_score = mean(Inspection.Score, na.rm = TRUE),
    failure_rate = mean(failed, na.rm = TRUE)
  )
# Count inspections by result type
result_counts <- inspection_clean |>
  count(result_group)

# Histogram of inspection scores
ggplot(inspection_clean, aes(x = Inspection.Score)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Inspection Scores",
    x = "Inspection Score",
    y = "Count"
  )
summary_stats
```

```{r}
# Bar chart of inspection outcomes
ggplot(result_counts, aes(x = result_group, y = n, fill = result_group)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Satisfactory" = "darkgreen",
    "Unsatisfactory" = "red",
    "Inspection Issues" = "orange",
    "Administrative" = "gray"
  )) +
  labs(
    title = "Inspection Outcomes (Grouped)",
    x = "Result Category",
    y = "Number of Inspections"
  )
```


# Where Problems Occur

This page explores the geographic distribution of food inspection results across King County.

We focus on **failure rates** (proportion of unsatisfactory inspections) to identify 
areas where food safety problems are more common.

TODO: Overlay points on a map
```{r}
#| warning: false
#| message: false

# Ensure we only use valid inspections
inspection_valid <- inspection_clean |>
  filter(result_group %in% c("Satisfactory", "Unsatisfactory"))

# Aggregate by ZIP code
zip_summary <- inspection_valid |>
  group_by("Zip Code") |>
  summarise(
    n_inspections = n(),
    failure_rate = mean(failed, na.rm = TRUE),
    avg_score = mean(Inspection.Score, na.rm = TRUE),
    latitude = mean(Latitude, na.rm = TRUE),
    longitude = mean(Longitude, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_inspections >= 30)  # remove very small sample sizes to help clean
```

```{r}
#| warning: false
#| message: false

# Map of failure rates
ggplot(inspection_valid, aes(x = Longitude, y = Latitude)) +
  geom_point(aes(color = failed), alpha = 0.3, size = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "green")) +
  labs(
    title = "Inspection Outcomes by Location",
    x = "Longitude",
    y = "Latitude",
    color = "Failed Inspection"
  )
```

